Fail Faster

### 2025-03-13 paper/paper.tex
CP 4.2

In Section~\ref{section:motiv}, we noticed that generator combinators like \texttt{weighted\_union}
must allocate lists in the hot path of the generator. Even though these lists are often small ---
usually at most a few dozen elements in practice --- each allocation takes us closer to the next garbage collection.

This is an ideal opportunity to exercise another use of staging: compile-time specialization.
Because we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
can generate \emph{different code} depending on the number of generators that are being unioned together.
If we use weighted union on a list (known at compile time) of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
we can emit code that never has to create a list of three elements at run time.

@@

Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
generators and weights. The weights themselves might only be known at run time --- it is
common to use the current size parameter as a weight, for instance ---
so they are \texttt{code}s.
\texttt{Gen.weighted\_union} begins by computing \texttt{sum\_code}, an \texttt{int code}
that is the sum of the weights. Note that this happens at compile time: we fold over a list
known at compile time to produce another code value. 
We then call \texttt{SR.int} to sample a random number \texttt{r} between \texttt{0} and the sum.
Finally, we splice in the result of calling the helper function \texttt{pick}. \texttt{pick}
produces a tree of \texttt{if}s by again traversing the list of generators at compile time.
This tree of \texttt{ifs} ``searches'' for the genrator corresponding to the sampled value \texttt{r}, and then runs it.

>>>>

In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
must allocate lists in the hot path of the generator. Even though these lists are often small ---
usually at most a few dozen elements in practice --- each allocation takes us closer to the next garbage collection.

This is an ideal opportunity to exercise another use of staging: compile-time specialization.
Since we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
can generate \emph{different code} depending on the number of generators in the union.
If we use weighted union on a compile-time list of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
we can emit code that picks between the generators without realizing the list at
run time.

@@

Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
generators and weights. The weights themselves might only be known at run time --- it is
common to use the current size parameter as a weight, for instance ---
so they are \texttt{code}s.
\texttt{Gen.weighted\_union} begins by computing \texttt{sum\_code}, an \texttt{int code}
that is the sum of the weights. Note that this happens at compile time: we fold over a list
known at compile time to produce another code value. 
We then call \texttt{SR.int} to sample a random number \texttt{r} between \texttt{0} and the sum.
Finally, we splice in the result of calling the helper function \texttt{pick}. \texttt{pick}
produces a tree of \texttt{if}s by again traversing the list of generators at compile time.
This tree of \texttt{ifs} ``searches'' for the generator corresponding to the
sampled value \texttt{r}, and then runs it.
\hg{I'm not sure this helps me. I think I'd prefer a less detailed explanation
of the code that conveys the intuition and let the reader actually read the code
if they want to know specifics. As it stands the explanation is just too dense
for me to process}
